Search Strategies for Functional Logic Programming

### 2012-01-18 papers/atps2012/paper.tex
ATPS: Benchmark comments improved

\author{
Michael Hanus
\quad
Bj{\"o}rn Peem{\"o}ller
\quad
Fabian Reck \\
\\
Institut f\"ur Informatik, CAU Kiel, D-24098 Kiel, Germany \\
\texttt{\{mh|bjp|fre\}@informatik.uni-kiel.de}
}
\title{Search Strategies for Functional Logic Programming}
\begin{document}
\maketitle

@@

In this section we evaluate the practical behavior
of the various search strategies discussed so far.
Since they are only supported by the Curry implementation KiCS2,
we use only this system for our evaluation.
A general comparison of KiCS2 and other Curry implementations
can be found in \cite{BrasselHanusPeemoellerReck11}.
Since KiCS2 compiles Curry programs into Haskell programs,
we used the Glasgow Haskell Compiler
(GHC 7.0.4, option -O2) to compile and execute
the generated target programs.
All benchmarks were executed on a 32bit Linux machine
running Ubuntu ???? with an Intel Core 2 Duo (???GHz) processor and 4 GiB RAM.
The timings were performed with the time command measuring the
execution time (in seconds) of a compiled executable for each benchmark.
Due to the lack of precise measurements of the space behavior,
we measured only the execution times where
``oom'' denotes a memory overflow in a computation.

@@

The table entries in Figure~\ref{fig:bench-all-solutions}
contain the run times in seconds to compute all values of the initial
expression.
As one can see, the top-level depth-first search is the most efficient
search strategy.
The encapsulated version of depth-first search (eDFS)
with the explicit construction of the search tree
causes some overhead, since the search strategy
is encoded in the source program rather than in the run-time system
as in the top-level search (DFS).
The benchmarks also show that
complete search strategies, like ``eBFS'' and ``eIDS(*2)''
are viable alternatives to the incomplete depth-first search strategy.
Their overhead (sometimes they are even faster than ``eDFS'',
but they are always slower than top-level search)
is acceptable taking into account the fact
that one need not to reason about the details of exploring the
search space. The behavior of iterative deepening is largely
influenced by the parameters.
A constant increment of the depth bound causes a big overhead
compared to doubling the depth bound in each iteration,
in particular, when the search tree is slim as in ``Last''.
This is even worse for top-level iterative deepening
which recomputes the initial expression in each iteration,
see ``IDS(+1)''.

Breadth-first search seems a good strategy
taking into account the size of memory provided by modern computers.
However, there are extreme cases when the search tree is wide
where iterative deepening with recomputation is the single
strategy which is able to compute a result.
This is shown in Figure~\ref{fig:bench-first-solution}
that contains benchmarks to compute only a first value
of an expression.

\begin{figure}[ht]
\centering
\begin{tabular}{|@{~~}l@{~~}|*{7}{@{~~}r@{~~}|}}
\hline
Program  &   DFS  & IDS(+1)& IDS(*2)&  eDFS  &  eBFS  &eIDS(+1)&eIDS(*2)\\
\hline
PermSort &  12.57 & 804.82 &  32.09 &  34.75 &  67.02 &  70.47 &  46.66 \\
Last     &   0.10 & 661.92 &   0.24 &   0.35 &   0.19 &   6.57 &   0.39 \\
Half     &   0.34 &  72.64 &   0.58 &   0.69 &   0.59 &   0.78 &   0.78 \\
Graph    &   0.00 &   0.10 &   0.04 &   0.00 &   0.04 &   0.04 &   0.03 \\
HorseMan &   2.06 & 636.62 &   6.64 &   2.52 &   2.60 &   2.61 &   2.57 \\
MAC      &   0.14 &  38.75 &   0.94 &   0.94 &   2.81 &   2.75 &   2.79 \\
Queens   &   1.40 & 370.26 &  33.65 &   2.44 &   8.95 &   8.86 &   3.04 \\
NDNums   &   oom  &  47.89 &   0.14 &    oom &   oom  &    oom &   0.47 \\
\hline
\end{tabular}
\caption{Benchmarks: computing a single value}
\label{fig:bench-first-solution}
\end{figure}

>>>>

\author{
Michael Hanus
\quad
Bj{\"o}rn Peem{\"o}ller
\quad
Fabian Reck \\
\\
Institut f\"ur Informatik, CAU Kiel, D-24098 Kiel, Germany \\
\texttt{\{mh|bjp|fre\}@informatik.uni-kiel.de}
}
\title{Search Strategies for Functional Logic Programming}
\begin{document}
\pagestyle{plain} % remove it in the camera-ready version
\maketitle

@@

In this section we evaluate the practical behavior
of the various search strategies discussed so far.
Since they are only supported by the Curry implementation KiCS2,
we use only this system for our evaluation.
A general comparison of KiCS2 and other Curry implementations
can be found in \cite{BrasselHanusPeemoellerReck11}.
Since KiCS2 compiles Curry programs into Haskell programs,
we used the Glasgow Haskell Compiler
(GHC 7.0.4, option -O2) to compile and execute
the generated target programs.
All benchmarks were executed on a 32bit Linux machine
running Ubuntu ???? with an Intel Core 2 Duo (???GHz) processor and 4 GiB RAM.
The timings were performed with the time command measuring the
execution time (in seconds) of a compiled executable for each benchmark.
Due to the lack of precise measurements of the space behavior,
we measured only the execution times.

@@

The table entries in Figure~\ref{fig:bench-all-solutions}
contain the run times in seconds to compute all values of the initial
expression.
As one can see, the top-level depth-first search is the most efficient
search strategy.
The encapsulated version of depth-first search (eDFS)
with the explicit construction of the search tree
causes some overhead, since the search strategy
is encoded in the source program rather than in the run-time system
as in the top-level search (DFS).
The benchmarks also show that
complete search strategies, like ``eBFS'' and ``eIDS(*2)''
are viable alternatives to the incomplete depth-first search strategy.
Their overhead (sometimes they are even faster than ``eDFS'',
but they are always slower than top-level search)
is acceptable taking into account the fact
that one need not to reason about the details of exploring the
search space. The behavior of iterative deepening is largely
influenced by its parameterization.
A constant increment of the depth bound causes a big overhead
compared to doubling the depth bound in each iteration,
in particular, when the search tree is slim as in ``Last''.
This is even worse for top-level iterative deepening
which recomputes the initial expression in each iteration,
see ``IDS(+1)''.

\begin{figure}[t]
\centering
\begin{tabular}{|@{~~}l@{~~}|*{7}{@{~~}r@{~~}|}}
\hline
Program  &   DFS  & IDS(+1)& IDS(*2)&  eDFS  &  eBFS  &eIDS(+1)&eIDS(*2)\\
\hline
PermSort &  12.57 & 804.82 &  32.09 &  34.75 &  67.02 &  70.47 &  46.66 \\
Last     &   0.10 & 661.92 &   0.24 &   0.35 &   0.19 &   6.57 &   0.39 \\
Half     &   0.34 &  72.64 &   0.58 &   0.69 &   0.59 &   0.78 &   0.78 \\
Graph    &   0.00 &   0.10 &   0.04 &   0.00 &   0.04 &   0.04 &   0.03 \\
HorseMan &   2.06 & 636.62 &   6.64 &   2.52 &   2.60 &   2.61 &   2.57 \\
MAC      &   0.14 &  38.75 &   0.94 &   0.94 &   2.81 &   2.75 &   2.79 \\
Queens   &   1.40 & 370.26 &  33.65 &   2.44 &   8.95 &   8.86 &   3.04 \\
NDNums   &   oom  &  47.89 &   0.14 &    oom &   oom  &    oom &   0.47 \\
\hline
\end{tabular}
\caption{Benchmarks: computing a single value}
\label{fig:bench-first-solution}
\end{figure}

The benchmarks indicate that breadth-first search seems to be a good strategy
taking into account the size of memory provided by modern computers.
However, if the search tree is wide (i.e., the nodes in each level
increases with the depth of the tree),
then iterative deepening is superior to breadth-first search.
To examine such a case, we added a benchmark ``NDNums''
which defines a non-deterministic operation
with a high branching factor that returns all inreasing numbers
\begin{curry}
g n = g (n+1) ? n ? g (n+1)
\end{curry}
and solve the equation \ccode{g 0 =:= 29}.
Obviously, the computation of all values will never terminate.
Therefore, we executed the benchmarks to compute
only a first value of an expression via different search strategies.
Figure~\ref{fig:bench-first-solution} contains the
corresponding results where
``oom'' denotes a memory overflow in a computation.
As one can see, iterative deepening is the only search strategy
that is able to find a solution in all examples.
